---
title: "Intermediate Elective 1"
date: "10 May 2026"
author: "Erick Morales Oyola"
format: pdf
editor: source
execute:
  message: false
  warning: false
---

# Setup 

## Packages

```{r}
#| label: packages

#Loading general use and cleaning packages
library(tidyverse)
library(here)
library(janitor)

#Loading date package
library(lubridate)

#Loading package for wrapping long axis labels
library(scales)

#Loading package for arranging plots together
library(patchwork)

#Loading packages for more customized figure styling
library(colorspace)
library(ggtext)
library(ggdist)
```

## Data

```{r}
#| label: data

#Reading in water quality survey metadata
metadata <- read_csv(here("data", "NCOS_YSI_Water_Quality_Monitoring_0.csv"),
                     show_col_types = FALSE)

#Reading in water quality parameter measurements
water_parameters <- read_csv(here("data", "YSI_Data_Begin_1.csv"),
                             show_col_types = FALSE)
```

## Cleaning 

```{r}
#Creating a cleaned metadata dataset

metadata_clean <- metadata |>
  #Cleaning column names
  clean_names() |>
  #Excluding irrelevant columns
  select(!c(object_id, creator, editor)) |>
  #Making sure monitoring date is in datetime format
  mutate(monitoring_date = lubridate::mdy_hms(monitoring_date)) |>
  #Renaming monitoring date column to clarify that it includes time
  rename(monitoring_datetime = monitoring_date) |>
  #Creating a column for site name
  mutate(site_full = case_when(
    site_name == "east_channel" ~ "East Channel",
    site_name == "phelps_bridge" ~ "Phelps Bridge",
    site_name == "venoco_bridge" ~ "Venoco Bridge",
    site_name == "other" ~ "Other"
  ))
```

```{r}
#| label: cleaning-water-parameters

#Creating a cleaned water parameters dataset
water_parameters_clean <- water_parameters |>
  #Cleaning column names
  clean_names() |>
  #Excluding unnecessary columns
  select(!c(creator, editor)) |>
  #Renaming depth code to elevation code
  rename(elevation_code = depth_code) |>
  #Replacing incorrect elevation code from 2021-07-16
  mutate(elevation_code = case_when(
    elevation_code == 10 ~ 1,
    .default = elevation_code
  )) |>
  #Setting elevation code as a factor
  mutate(elevation_code = as_factor(elevation_code),
         elevation_code = fct_relevel(elevation_code,
                                      "0", "1", "2", "3", "4", "5", "6"))
```

```{r}
#| label: joining-water-quality-data

#Joining metadata and water parameter measurements
water_quality <- full_join(
  #Using metadata as the left dataframe
  x = metadata_clean,
  #Using water parameters as the right dataframe
  y = water_parameters_clean,
  #Joining survey records to their matching parameter measurements
  by = c("global_id" = "parent_global_id")
) |>
  #Selecting relevant columns for the elective visualization
  select(global_id,
         monitoring_datetime,
         site_name,
         site_full,
         water_surface_elevation,
         elevation_code,
         do_mg_l,
         do_percent_sat,
         conductivity_u_s_cm,
         salinity_ppt,
         temperature_c) |>
  #Extracting the date from monitoring datetime
  mutate(date = lubridate::date(monitoring_datetime)) |>
  #Moving date next to monitoring datetime for readability
  relocate(date, .after = monitoring_datetime) |>
  #Removing the original datetime column
  select(!monitoring_datetime)
```

```{r}
#| label: cleaning-elective-data

#Creating a cleaned water quality dataset for the elective figure
water_quality_clean <- water_quality |>
  #Filtering out other sites so the figure focuses on the three main sites
  filter(site_full != "Other") |>
  #Filtering out extreme salinity outliers
  filter(salinity_ppt < 1410) |>
  #Filtering to the main elevation codes used in class
  filter(elevation_code %in% c("0", "1", "2")) |>
  #Removing rows without the main variables needed for the figure
  filter(!is.na(temperature_c),
         !is.na(do_mg_l),
         !is.na(salinity_ppt)) |>
  #Ordering sites from lower to higher salinity
  mutate(site_full = fct_relevel(site_full,
                                 "Phelps Bridge",
                                 "East Channel",
                                 "Venoco Bridge"))

#Displaying 10 random rows from the cleaned elective dataset
water_quality_clean |>
  slice_sample(n = 10)
```

```{r}
#| label: site-summary-data

#Creating site-level summary values for the scatterplot
site_summary <- water_quality_clean |>
  #Grouping by sampling site
  group_by(site_full) |>
  #Calculating median and standard deviation values for each site
  summarize(med_temperature = median(temperature_c, na.rm = TRUE),
            med_oxygen = median(do_mg_l, na.rm = TRUE),
            med_salinity = median(salinity_ppt, na.rm = TRUE),
            sd_temperature = sd(temperature_c, na.rm = TRUE),
            sd_oxygen = sd(do_mg_l, na.rm = TRUE),
            n = n()) |>
  #Ungrouping the dataframe
  ungroup()

#Displaying the site-level summary data
site_summary
```


# Problems 

## 1. Explain your inspiration 

For my inspiration, I chose Cédric Scherer’s “Bill Dimensions of Brush-tailed Penguins.” I really like how the figure combines individual observations with group-level summaries, using color, point size, labels, and a second distribution panel to make the patterns feel informative while being visually polished.

This inspiration makes sense for the water quality dataset because the data also includes multiple continuous variables that can be explored together. Instead of comparing penguin bill length, bill depth, and body mass by species, I can compare water temperature, dissolved oxygen, and salinity across three sites at North Campus Open Space.

In my visualization, water temperature will be shown on the x-axis, dissolved oxygen will be shown on the y-axis, salinity will be represented by point size, and site will be represented by color. I also plan to include site-level summary markers or error bars to show median conditions per each site, following the same general idea as the penguin figure’s species-level summaries.

## 2. Plan your figure

!! DONT FORGET TO INSERT SKETCH !!!!!!!!! PLEASEEEEE

For my figure, I plan to adapt the structure of the penguin bill-dimensions visualization into a water quality figure. The top panel will be a scatterplot showing water temperature on the x-axis and dissolved oxygen on the y-axis, with point color representing site and point size representing salinity. I also plan to include larger site-level summary points and error bars to show median conditions for each site.

The bottom panel will show the distribution of salinity across sites, similar to the distribution panel in the inspiration figure. This will help connect individual water quality measurements to broader site-level differences in salinity.

## 3. Code your figure

```{r}
#| label: 3-figure-color-palette

#Creating a color palette inspired by the bright colors in the penguin figure
site_palette <- c(
  "Phelps Bridge" = "#159090",
  "East Channel" = "#FF8C00",
  "Venoco Bridge" = "#A034F0"
)
```

```{r}
#| label: 3-figure-data

#Creating a focused dataset for the final elective figure
elective_figure_data <- water_quality_clean |>
  #Creating year and month columns from the date column
  mutate(year = year(date),
         month = month(date)) |>
  #Creating water year so the figure matches the class workflow
  mutate(wy = case_when(
    month >= 10 ~ year + 1,
    .default = year
  )) |>
  #Keeping only water year 2024 to make the figure more focused and readable
  filter(wy == "2024") |>
  #Keeping only the three main NCOS sites
  filter(site_full %in% c("Phelps Bridge",
                          "East Channel",
                          "Venoco Bridge")) |>
  #Removing missing values in the main variables used for plotting
  filter(!is.na(temperature_c),
         !is.na(do_mg_l),
         !is.na(salinity_ppt))

#Creating site-level summary values for the scatterplot
elective_site_summary <- elective_figure_data |>
  #Grouping by site
  group_by(site_full) |>
  #Calculating median and standard deviation values for each site
  summarize(med_temperature = median(temperature_c, na.rm = TRUE),
            med_oxygen = median(do_mg_l, na.rm = TRUE),
            med_salinity = median(salinity_ppt, na.rm = TRUE),
            sd_temperature = sd(temperature_c, na.rm = TRUE),
            sd_oxygen = sd(do_mg_l, na.rm = TRUE),
            n = n()) |>
  #Ungrouping the dataframe
  ungroup() |>
  #Creating slightly adjusted oxygen positions so overlapping 
  #error bars are readable
  mutate(plot_med_oxygen = case_when(
    site_full == "Phelps Bridge" ~ med_oxygen - 0.25,
    site_full == "East Channel" ~ med_oxygen + 0.35,
    site_full == "Venoco Bridge" ~ med_oxygen - 0.35
  ))

#Displaying the summary data used for the group-level figure layers
elective_site_summary
```


```{r}
#| label: 3-scatterplot
#| fig-width: 9
#| fig-height: 6

#Creating the top panel of the figure
water_quality_scatter <- ggplot(data = elective_figure_data,
                                #Mapping temperature, dissolved oxygen, 
                                #site, and salinity
                                mapping = aes(x = temperature_c,
                                              y = do_mg_l,
                                              color = site_full,
                                              size = salinity_ppt)) +
  #Adding individual water quality measurements as opaque points
  geom_point(alpha = 0.5) +
  #Adding vertical error bars to show dissolved oxygen variation 
  #around each site median
  geom_errorbar(data = elective_site_summary,
              mapping = aes(x = med_temperature,
                            ymin = plot_med_oxygen - sd_oxygen,
                            ymax = plot_med_oxygen + sd_oxygen,
                            color = site_full),
              inherit.aes = FALSE,
              width = 0.35,
              linewidth = 0.8,
              show.legend = FALSE) +
  #Adding horizontal error bars to show temperature variation 
  #around each site median
  geom_errorbar(data = elective_site_summary,
              mapping = aes(y = plot_med_oxygen,
                            xmin = med_temperature - sd_temperature,
                            xmax = med_temperature + sd_temperature,
                            color = site_full),
              inherit.aes = FALSE,
              orientation = "y",
              width = 0.25,
              linewidth = 0.8,
              show.legend = FALSE) +
  #Adding larger median points for each site
  geom_point(data = elective_site_summary,
           mapping = aes(x = med_temperature,
                         y = plot_med_oxygen,
                         fill = site_full,
                         size = med_salinity),
           inherit.aes = FALSE,
           shape = 21,
           color = "white",
           stroke = 0.9,
           show.legend = FALSE) +
  #Using the custom site colors
  scale_color_manual(values = site_palette) +
  #Using matching fill colors for the median points
  scale_fill_manual(values = site_palette) +
  #Scaling point size to show salinity without overwhelming the plot
  scale_size_continuous(name = "Salinity (ppt)",
                        range = c(0.6, 5),
                        breaks = c(0, 20, 40, 60)) +
  #Adding clean axis labels, legend labels, title, and subtitle
  labs(x = "Water temperature (°C)",
       y = "Dissolved oxygen (mg/L)",
       color = "Site",
       title = "Water quality dimensions across NCOS sampling sites",
       subtitle = "A. Water temperature and dissolved oxygen in water year 2024") +
  #Using a clean non-default theme
  theme_bw() +
  #Customizing the plot theme for a more polished final figure
  theme(legend.position = "right",
        plot.title = element_text(face = "bold", size = 15),
        plot.subtitle = element_text(face = "italic", size = 10),
        panel.grid.minor = element_blank())

#Displaying the scatterplot
water_quality_scatter
```
## 4. Describe your process

For this elective, I used Cédric Scherer’s penguin bill-dimensions visualization as a guide for my figure’s structure. I liked how the original figure combined individual observations, group-level summaries, error bars, color, and point size, and I adapted those ideas to display water quality variables from three sites at NCOS.

One challenge was figuring out how to translate the penguin variables into meaningful water quality variables without making the figure too crowded. At first, the scatterplot had too many overlapping points and labels, so I narrowed the data to water year 2024, reduced the point opacity, removed the labels, and relied more on the legend to keep the figure readable.

This visualization differs from the figures I made in class and previous assignments because I designed it as a polished graphic rather than a simpler exploratory plot. In class, most figures focused on showing one relationship clearly, while this figure combines several visual elements at once, including individual measurements, site-level medians, error bars, and an additional variable as point size.
